P3064R0: How to Avoid OOTA Without Really Trying

### 2024-05-04 ootavacuous.tex
Tokyo is done, so post-Tokyo it is!
Reported-by: Alan Stern <stern@rowland.harvard.edu>
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>
Upgrade from "D" to "P" for submission
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>
Add more compelling examples to "Inventing Atomic Loads"
Also add subsections.

Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

\begin{document}
\title{D3064R0: How to Avoid OOTA Without Really Trying}

@@

\author{
Alan Stern\\{\small \email{stern@rowland.harvard.edu}} \and
Paul E.~McKenney\\{\small \email{paulmck@kernel.org}} \and
Michael Wong\\{\small \email{fraggamuffin@gmail.com}} \and
Maged Michael\\{\small \email{maged.michael@gmail.com}}
}
\date{April 5, 2024 (Post-Pre-Tokyo)}
\maketitle{}

@@

@@

@@

There are situations where inventing loads from non-atomic objects
is customary, for example, when hoisting a load out of a loop.
In this case, compilers might hoist the load from \co{x} out of
the loop:
\begin{quote}
\begin{verbatim}
do_something_with(x);
for (i = 0; i < limit; i++)
    y[i] += x;
\end{verbatim}
\end{quote}
A compiler might act as if the program had instead been this:
\begin{quote}
\begin{verbatim}
r1 = x;
do_something_with(r1);
for (i = 0; i < limit; i++)
    y[i] += r1;
\end{verbatim}
\end{quote}
Note that passing \co{x} to \co{do_something_with()} is important
when \co{x} is not atomic,
because this allows the compiler to assume that inventing additional
accesses to \co{x} will not result in a data race.
Without this assumption, the above transformation might introduce a data
race when \co{limit} is less than or equal to zero.

@@

>>>>

\begin{document}
\title{P3064R0: How to Avoid OOTA Without Really Trying}

@@

\author{
Alan Stern\\{\small \email{stern@rowland.harvard.edu}} \and
Paul E.~McKenney\\{\small \email{paulmck@kernel.org}} \and
Michael Wong\\{\small \email{fraggamuffin@gmail.com}} \and
Maged Michael\\{\small \email{maged.michael@gmail.com}}
}
\date{April 5, 2024 (Post-Tokyo)}
\maketitle{}

@@

\subsubsection{Illustrative Atomic-Load Invention}
\label{app:Illustrative Atomic-Load Invention}

@@

\subsubsection{Illustrative Atomic-Load Use Invention}
\label{app:Illustrative Atomic-Load Use Invention}

@@

However, before making such a choice, we should look at other situations.

\subsubsection{Customary Non-Atomic-Load ``Invention''}
\label{app:Customary Non-Atomic-Load Invention}

There are situations where inventing loads from non-atomic objects
is customary, for example, when hoisting a load out of a loop.
In this case, compilers might hoist the load from \co{x} out of
the loop:
\begin{quote}
\begin{verbatim}
do_something_with(x);
for (i = 0; i < limit; i++)
    y[i] += x;
\end{verbatim}
\end{quote}
A compiler might act as if the program had instead been this:
\begin{quote}
\begin{verbatim}
r1 = x;
do_something_with(r1);
for (i = 0; i < limit; i++)
    y[i] += r1;
\end{verbatim}
\end{quote}
Note that passing \co{x} to \co{do_something_with()} is important
when \co{x} is not atomic,
because this allows the compiler to assume that inventing additional
accesses to \co{x} will not result in a data race.
Without this assumption, the above transformation might introduce a data
race when \co{limit} is less than or equal to zero.
Hence the use of quotation marks in the section heading:
Here the load is not actually invented, but instead fused with an
earlier load with no synchronization between the two loads.

@@

In addition, compiler writers should carefully consider the issues
raised by the following section:

\subsubsection{Other Cautions for Atomic-Load Invention}
\label{app:Other Cautions for Atomic-Load Invention}

Consider the code below:
\begin{quote}
\begin{verbatim}
int r0 = x.load(memory_order_relaxed);
int r1 = r0 * r0 + 2 * r0 + 1;
\end{verbatim}
\end{quote}
A developer might be forgiven for assuming that the value of \co{r1}
is guaranteed to be a perfect square, at least for sufficiently small
values of \co{x}.

But if the compiler is permitted to invent atomic loads, there is no such
guarantee, as can be seen from this transformed version of the above code:
\begin{quote}
\begin{verbatim}
int r0 = x.load(memory_order_relaxed);
int invented = x.load(memory_order_relaxed);
int r1 = r0 * r0 + 2 * invented + 1;
\end{verbatim}
\end{quote}
For example, if the load into \co{r0} happened when the value of \co{x}
was zero and the load into \co{invented} happened when the value of \co{x}
was ten, then the value of \co{r1} will be twenty, which is a perfectly
legitimate integer, but not a perfect square.

Similar issues can arise even when the compiler merely repurposes a
value from an atomic load, as can be seen from this example:
\begin{quote}
\begin{verbatim}
do_something(x.load(memory_order_relaxed));
int r0 = x.load(memory_order_relaxed);
int r1 = r0 * r0 + 2 * r0 + 1;
\end{verbatim}
\end{quote}
This again ensures that \co{r1} will be a perfect square given sufficiently
small values of \co{x}.
But is the compiler allowed to transform this example into the following?
\begin{quote}
\begin{verbatim}
repurposed = x.load(memory_order_relaxed);
do_something(repurposed);
int r0 = x.load(memory_order_relaxed);
int r1 = r0 * r0 + 2 * repurposed + 1;
\end{verbatim}
\end{quote}
Again, this can result in the value of \co{r1} being something other than
a perfect square.
%
But it would be just fine if the compiler fully fused the two loads from
\co{x}, for example, as follows:
\begin{quote}
\begin{verbatim}
r0 = x.load(memory_order_relaxed);
do_something(r0);
int r0 = x.load(memory_order_relaxed);
int r1 = r0 * r0 + 2 * r0 + 1;
\end{verbatim}
\end{quote}

Similar observations apply to atomic loads from pointers.
For example, consider a load from an atomic pointer and subsequent uses
of the pointed-to fields:
\begin{quote}
\begin{verbatim}
do_something(gp.load(memory_order_relaxed));
p = gp.load(memory_order_relaxed);
do_something_else(p->a, p->b,p->c);
\end{verbatim}
\end{quote}
Users are within their rights to assume that the three arguments to
\co{do_something_else()} will be taken from the same instance of the pointed-to
structure, at least assuming that the lifetime of that structure extends
across all of the above statements.
However, further assuming that \co{do_something()} is free of
synchronization, a compiler would be within its rights to fuse the two
loads from \co{gp}, for example, as follows:
\begin{quote}
\begin{verbatim}
p = gp.load(memory_order_relaxed);
do_something(p);
do_something_else(p->a, p->b,p->c);
\end{verbatim}
\end{quote}

It is imperative that users be able to rely on arithmetic and pointer accesses.
Therefore, compilers must refrain from inventing atomic loads and from repurposing
their values, but compilers are permitted to fuse adjacent atomic loads from the
same variable.
